Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-03-17 paper.tex
update

\citeauthor{van2012systematic} proposed Abstracting Abstract Machines (AAM) methodology which can used
for constructing abstract interpreters of higher-order functional languages from concrete abstract machines
~\cite{van2012systematic, van2010abstracting}.
Given a concrete small-step abstract machine, (e.g, the CESK machine, Krivine's machine, etc.),
by allocating continuations in the store and bounding both the value addresses and continuation
addresses to be finite, we obtain an abstract interpreter with a finite state space which
can be used for performing sound static analysis.
One can further instantiate different polyvariant control flow analyses by using different address allocators \cite{Gilray:2016:ACP:2951913.2951936}.

@@

Based on the refunctionalized AAM from the last section, we use the same @nd@ nondeterministic operator
but further extend it to a fixed-point cached evaluation in this section. \todo{extend this}

@@

\textbf{Cache.}
We will apply the fixed-point caching algorithm as described from
\citeauthor{darais2017abstracting}'s ADI paper.
Note that the latent assumption of the worklist algorithm in classical AAM
is if we have seen a state @s@, then it means we also have seen the successors
of state @s@. 
The caching algorithm we adopted here replay the same assumption but in a big-step
manner: if we have seen the configuration @c@, then it means we also have seen
the values that are evaluated by the configuration @c@.

@@

Otherwise, if we not hit the @out@ cache, we retrieve the values from @in@ cache, and update @out@ cache
with values from @in@ cache.  The retrieving from @in@ cache may return an empty set of values if 
there is no such mapping for the configuration from the previous iteration.
It actually first conservatively assumes that all computation can be diverged,
if not, we will update its values in the mapping after evaluation.
In terms of partial orders and lattices, it is also the case we start 
the Kleene iteration from the bottom of the lattice.

@@

\begin{lstlisting}
def analyze(e: Expr, env: Env = mtEnv, store: BStore = mtStore) = {
  def iter(cache: Cache): Ans = {
    val Ans(vss, new_cache) = aeval(e, env, store, mtTime, cache, {
      val initConfig = Config(e, env, store, mtTime)
      case Ans(vss, cache) => Ans(vss, cache.outUpdate(initConfig, vss))
    })
    if (new_cache.out == cache.out) { Ans(vss, new_cache) }
    else { iter(Cache(new_cache.out, new_cache.out)) }
  }
  iter(mtCache)
}
\end{lstlisting}

>>>>

\citeauthor{van2012systematic} proposed Abstracting Abstract Machines (AAM) methodology which can used
for constructing sound abstract interpreters of higher-order functional languages from concrete abstract machines
~\cite{van2012systematic, van2010abstracting}.
Given a concrete small-step abstract machine, (e.g, the CESK machine, Krivine's machine, etc.),
by allocating continuations in the store and bounding both the value addresses and continuation
addresses to be finite, we obtain an abstract interpreter with a finite state space which
can be used for performing sound static analysis.
One can further instantiate different polyvariant control flow analyses by using different address allocators \cite{Gilray:2016:ACP:2951913.2951936}.

@@

Based on the refunctionalized AAM from the last section, we use the same @nd@ nondeterministic operator
but further extend it to a fixed-point cached evaluation in this section. 
The fixed-point caching algorithm performs a sound over-approximation of all possible 
concrete reachable pathes and values, and also prevents non-termination when analyzing diverged programs.

@@

\textbf{Cache.}
Recall that the latent assumption of the worklist algorithm in classical AAM
is if we have seen a state @s@, then it means we also have seen the successors
of state @s@. 
The caching algorithm we adopted here replay the same assumption but in a big-step
manner: if we have seen the configuration @c@, then it means we also have seen
the values that are evaluated by the configuration @c@.
Here we will apply the fixed-point caching algorithm as described from
\citeauthor{darais2017abstracting}'s ADI paper \cite{darais2017abstracting}.

@@

Otherwise, if we not hit the @out@ cache, we retrieve the values from @in@ cache, and update @out@ cache
with values from @in@ cache.  The retrieving from @in@ cache may return an empty set of values if 
there is no such mapping for the configuration from the previous iteration.
It actually first conservatively assumes that all computation can be diverged,
if not, we will update its values in the mapping after evaluation.
In terms of partial orders and lattices, it is the case that starts
the Kleene iteration from the bottom of the lattice.

@@

\begin{lstlisting}
def analyze(e: Expr) = {
  def iter(cache: Cache): Ans = {
    val Ans(vss, new_cache) = aeval(e, Map[String, BAddr](), Store[BAddr, Storable](Map()), List(), cache, {
      val initConfig = Config(e, Map[String, BAddr](), Store[BAddr, Storable](Map()), List())
      case Ans(vss, cache) => Ans(vss, cache.outUpdate(initConfig, vss))
    })
    if (new_cache.out == cache.out) { Ans(vss, new_cache) }
    else { iter(Cache(new_cache.out, new_cache.out)) }
  }
  iter(mtCache)
}
\end{lstlisting}

Finally, the @analyze@ function as the entrance of the analysis does a looping iteration
to find the fixed-point on caches starting from an empty cache. 
If the @out@ cache of this iteration is equivalent to the @out@ cache from the last iteration,
then we have reached a fixed-point that over-approximates the concrete evaluations,
and thus can be returned.
Otherwise, a next iteration with the new @out@ cache will be activated.
Note that the continuation of the initial call to @aeval@ sets up the values for the 
initial configuration in the final cache.
